Lifespan analysis

### 2024-04-17 paper/manuscript.Rmd
ready for revision

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze following across the lifespan (*N* = 471 3- to 80-year-olds; children from a mid-sized German city; international remotely tested adults). We found a steep performance improvement during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze following in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we empirically tested how gaze following relates to vector following in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze following is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze following, individual differences, cognitive modeling, lifespan"
wordcount         : "X"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
classoption       : "man"

@@

# Introduction
Following the gaze of others is valuable for extracting information from the environment. It guides us to "informational hotspots" [@meltzoff2010social, p.1] and can be used to identify internal states such as intentions or emotions [@corkum1998origins; @pfeiffer2013gaze]. As one of the most fundamental social-cognitive abilities, gaze-following is an integral part of almost every form of social interaction, including communication, collaboration, cultural and social learning [@emery2000eyes; @rakoczy2022foundations; @frith2012mechanisms; @tomasello2003makes; @bohn2019pervasive; @moore2008development; @hessels2020how], and has been extensively studied in infancy [for review, see @delbianco2019developmental]. In the most commonly used paradigm [e.g., @astor2020social; @gredeback2010development; @byers-heinlein2021development; @ishikawa2022physiological], the experimenter looks directly at the infant before shifting their head and eyes to one of two objects. Infants’ looking times to the target or the proportion of choosing the target over the distractor are measured. Research in this tradition finds that infants as young as three to four months can follow the gaze of another agent [@delbianco2019developmental; @dentremont2000perceptual; @dentremont1997demonstration; @astor2021gaze].

Previous research has shown a refinement of gaze-following abilities in a child's first and second year of life [e.g., @astor2021gaze; @butterworth1991minds; @brooks2002importance]. At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412; @deak2000effects; @butterworth1991minds; @corkum1995development]. However, we do not know much about the developmental progression beyond these qualitative milestones. One possibility is that the ability to follow gaze does not improve beyond infancy. Yet, most, if not all, cognitive abilities continue to develop throughout childhood [e.g., @gathercole2004structure; @gredeback2010development]. It seems likely that children fine-tune their gaze following as they get older - presumably while using them in social interactions. To capture the development in gaze following beyond infancy, Study 1 was included participants from preschool to old age.

@@

Group-level analyses of gaze-following abilities (e.g., average age at which children as a group reach an above-chance performance) may mask individual differences between children. Measuring individual differences in basic aspects of social cognition is important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. Across the three studies, we measured gaze following continuously by using a task that is designed to capture individual-level variation [@prein2023tango]: the TANGO avoids floor and ceiling effects in children and adults and is thus particularly suited to examine how gaze following changes with age.

@@

The study was pre-registered prior data collection: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission in Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45). Data was collected between May 2021 and April 2023.

@@

(ref:figlab1) **Developmental trajectory of gaze following across the lifespan**. Grey dots show the mean level of imprecision (ie., absolute distance between the target's center and the participant's click) for each subject averaged across trials. The unit of imprecision is counted in target width, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Blue lines show 100 draws from the expectation of the posterior predictive distribution of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the locations of the most prominent changes according to our Bayesian change point analysis.

@@

```{r lifespan_results, include=FALSE}
lifespan_variation <- readRDS("../saves/lifespan_variation_all.rds")
lifespan_mcomparison <- readRDS("../saves/lifespan_mcomparison.rds") %>% mutate_if(is.numeric, round, 2)
lifespan_mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")
lifespan_beast <- readRDS("../saves/lifespan_ibeast.rds")
lifespan_cps <- readRDS("../saves/lifespan_ibeast_cps.rds") %>% round(2) %>% arrange(cp)

@@

High levels of variation pointed to substantial individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]). We found substantial evidence for a non-linear development in gaze following across the lifespan. The Gaussian process model was clearly preferred over the polynomial models due to the highest predictive accuracy according to the LOO ELPD estimates: elpd_diff between Gaussian process and cubic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mcubic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mcubic"]`); elpd_diff between Gaussian process and quadratic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mquadratic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mquadratic"]`); elpd_diff between Gaussian process and linear model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mlinear"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mlinear"]`); all in favor of the Gaussian process model. Moreover, the Gaussian Process model showed the greatest model weight (approximating `r lifespan_mcomparison$modelweight[lifespan_mcomparison$model == "mgaussianprocess"]`). For the imprecision in gaze following, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) indicated nonlinearity.

The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp_mode` major shifts in gaze following during the lifespan (MAP estimate with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp_mode + 1] * 100 %>% round(2)`% probability). The change points occurred at `r lifespan_cps$cp[1]` (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`]); `r lifespan_cps$cp[2]` (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`]); `r lifespan_cps$cp[3]` (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`]); `r lifespan_cps$cp[4]` (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`]); `r lifespan_cps$cp[5]` (95% CrI [`r lifespan_cps$lowerCrI[5]`; `r lifespan_cps$upperCrI[5]`]); and `r lifespan_cps$cp[6]` years (95% CrI [`r lifespan_cps$lowerCrI[6]`; `r lifespan_cps$upperCrI[6]`]). In short: we found a rapid initial improvement in gaze following in early childhood, followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward old age.

@@

Previous studies found that already four-month-olds demonstrate basic gaze-following abilities [@delbianco2019developmental]. In our sample, three-year-olds were still rather imprecise in their gaze following ability (average imprecision was approx. two target widths). How can we explain this divergence? First, we used subtle eye movements as cues. Many existing studies let the agents move eye and head in parallel [@behne2005oneyearolds; @povinelli1997exploitation], establishing a confound with the more salient head movement. Relying exclusively on eye movements might be more difficult for children than presenting them with a combined eye and head orientation [@carpenter1998social]. @silverstein2021infants used a similar manipulation of gaze cues without head rotation and found that 6- to 18-month-olds were around or just above chance for gaze-following. The authors argue that infants might fixate on another's face most of the time, while eye movement alone might not be strong enough to guide their attention. Furthermore, our study required participants to (1) precisely follow an agent's gaze, (2) interpret this as a cue, and (3) use the cue to guide their own behavior. It is conceivable that three-year-olds followed the agent's gaze but did not translate this into precise, active behavior. @moll2013primacy argue that social forms of perspective-taking evolve prior to visual perspective-taking, which only emerges within the third year of life. Young children might simply not be interested in a differential, spatial representation of the surrounding objects. Taken together, this might explain why our younger participants located the agent's gaze rather imprecisely.

@@

Our lifespan study showed that gaze following develops throughout childhood, and variation between individuals appears in all age groups. The TANGO has previously been shown to reliably capture inter-individual differences in gaze following [@prein2023tango]. The variation between participants was thus likely genuine and not due to random noise. In Study 2, we aimed to understand the developmental change and individual differences on a process level. We present a theory of gaze following that explains how participants process the available gaze information and trace a line-of-sight to identify the agent's focus. We formalized this inference process in a computational cognitive model that replicates a schematic representation of how participants make inferences in the task’s context (i.e., model of the task and not the data).

@@

[^6]: In an exploratory analysis, we simulated data for two more alternatives: a line-of-sight tracing model that assumed no inferential noise in the participants' gaze-following ability, and a model building up on the line-of-sight tracing with added motor noise. Please note that these did not predict a U-shape pattern, since all target locations would be influenced by the motor noise equally. For further details, see Supplements.

@@

First, we experimentally isolated the vector following component of the TANGO. We designed a new non-social vector following task that shared all crucial design features of the TANGO. Second, we assessed children's social-cognitive abilities by administering a ToM task battery, comprising four tasks from the ToM scale by Wellman and Liu [@wellman2004scaling] and two additional perspective-taking tasks [@flavell1981development; @flavell1981younga]. We reasoned that the TANGO shares task demands with the non-social vector following task while it shares its social context with the ToM tasks. We could therefore imagine both an absence or presence of relationship betweenn gaze following and ToM. As stated in our pre-registration, we further assessed whether the two perspective-taking tasks related to gaze following. Our reasoning was that similar underlying mechanisms might be needed to solve these tasks since they require participants to take into account another person's point of view.

@@

We administered four tasks from the @wellman2004scaling ToM scale (see Supplements for further detail). We excluded three tasks: the Diverse Desires task to avoid ceiling effects; and both tasks involving emotions (Belief Emotion and Real-Apparent Emotion), as we aimed at assessing the "cold, cognitive" (vs. "emotional") aspects of social cognition. We added two perspective-taking level-2 tasks [@flavell1981development; @flavell1981younga; where children were asked whether a turtle appeared to be on its back or feet / a worm lay on a red or blue blanket from the experimenter's point of view] with the aim of increasing the variability we can capture between individuals, and since we hypothesized that perspective-taking would rely on similar mechanisms than gaze following, both relying on another's person egocentric frame of reference.

@@

(ref:figlab4) **Components of gaze following.** (A) Study procedures. Top: TANGO (i.e., gaze following; social vector following). Bottom: Magnet (i.e., non-social vector following). The left-hand side shows screenshots of the familiarization phase; right-hand side shows screenshots of the test phase. For illustrative purposes, transluscent targets are shown in the test phase. Participants cannot see them. (B) Correlations between gaze following, non-social vector following, ToM, and perspective-taking. Dots show the correlation coefficients, error bars represent 95% CIs. Please note that the ToM score acts as an umbrella term and also includes two perspective-taking tasks. (C) Influence of age, perspective-taking and non-social vector following on gaze following. The graph shows the posterior distributions for the respective predictor. Black dots represent means, thicker black lines 80% CrI, and thinner black lines 95% CrI.

@@

In this paper, we have focused on studying variation across ages and individuals. However, our findings rely on participants from Western, Educated, Industrialized, Rich, and Democratic (WEIRD) backgrounds [@henrich2010weirdesta]. Cultural variation has been found in many foundational aspects of cognition and socialization: for example, in parent-child interaction and communication [@nielsen2017persistent]. First evidence suggests that cultural variation in face-to-face interactions does not influence infants' gaze shifts [@hernik2019infanta]. While we cannot generalize our here-reported developmental trajectory to different socio-environmental settings, we predict that our presented process-model of gaze following holds true across communities. Analyzing cross-cultural variation and checking for predicted "signature patterns" in the data will inform our modeling work and theory building in further detail [@amir2020crosscultural].

In Study 1, we recruited older participants online which might have selected a particular subgroup of this age range. Seventy-year-olds who have working WIFI connections, know how to use a computer, and are registered on *Prolific* might not be representative of their age group. We can imagine that results from a more diverse, in-person data collection might show different developmental trajectories toward old age.

@@

In three studies, we have illuminated the lifelong development of precisely estimating another's gaze direction, and the psychophysical process behind this. We have shown that gaze following continues to develop beyond infancy, and that individuals differ in their precision levels to localize the gaze direction of an agent. Our proposed process-level theory of gaze following modeled individual differences in precision as varying levels of uncertainty in the estimated gaze vectors. Consequently, we found that imprecision in gaze following relates to non-social vector following, as proposed by the model. Additionally, gaze following was linked to visual perspective-taking but no other aspects of ToM. The present research shows how precise and reliable measures and process models jointly inform each other and lead to a more comprehensive understanding of the psychological phenomenon in question.

>>>>

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze following across the lifespan (*N* = 471 3- to 80-year-olds; children from a mid-sized German city; international remotely tested adults). We found a steep performance improvement during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze following in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we empirically tested how gaze following relates to vector following in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze following is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze following, individual differences, cognitive modeling, lifespan"
wordcount         : "10469"
floatsintext      : yes
figurelist        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
classoption       : "man"

@@

# Introduction
Following the gaze of others is valuable for extracting information from the environment. It guides us to "informational hotspots" [@meltzoff2010social, p.1] and can be used to identify internal states such as intentions or emotions [@corkum1998origins; @pfeiffer2013gaze]. As one of the most fundamental social-cognitive abilities, gaze following is an integral part of almost every form of social interaction, including communication, collaboration, cultural and social learning [@emery2000eyes; @rakoczy2022foundations; @frith2012mechanisms; @tomasello2003makes; @bohn2019pervasive; @moore2008development; @hessels2020how], and has been extensively studied in infancy [for review, see @delbianco2019developmental]. In the most commonly used paradigm [e.g., @astor2020social; @gredeback2010development; @byers-heinlein2021development; @ishikawa2022physiological], the experimenter looks directly at the infant before shifting their head and eyes to one of two objects. Infants’ looking times to the target or the proportion of choosing the target over the distractor are measured. Research in this tradition finds that infants as young as three to four months can follow the gaze of another agent [@delbianco2019developmental; @dentremont2000perceptual; @dentremont1997demonstration; @astor2021gaze].

Previous research has shown a refinement of gaze following abilities in a child's first and second year of life [e.g., @astor2021gaze; @butterworth1991minds; @brooks2002importance]. At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412; @deak2000effects; @butterworth1991minds; @corkum1995development]. However, we do not know much about the developmental progression beyond these qualitative milestones. One possibility is that the ability to follow gaze does not improve beyond infancy. Yet, most, if not all, cognitive abilities continue to develop throughout childhood [e.g., @gathercole2004structure; @gredeback2010development]. It seems likely that children fine-tune their gaze following as they get older - presumably while using them in social interactions. To capture the development in gaze following beyond infancy, Study 1 included participants from preschool to old age.

@@

Group-level analyses of gaze following abilities (e.g., average age at which children as a group reach an above-chance performance) may mask individual differences between children. Measuring individual differences in basic aspects of social cognition is important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. Across the three studies, we measured gaze following continuously by using a task that is designed to capture individual-level variation [@prein2023tango]: the TANGO (Task for Assessing iNdividual differences in Gaze understanding - Open) avoids floor and ceiling effects in children and adults and is thus particularly suited to examine how gaze following changes with age.

@@

The study was pre-registered prior to data collection: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission in Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45). Data was collected between May 2021 and April 2023.

@@

(ref:figlab1) **Developmental trajectory of gaze following across the lifespan**. Grey dots show the mean level of imprecision (ie., absolute distance between the target's center and the participant's click) for each subject averaged across trials. The unit of imprecision is counted in target width, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Blue lines show 100 draws from the expectation of the posterior predictive distribution of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the locations of the changes with the highest posterior probability according to our Bayesian change point analysis.

@@

```{r lifespan_results, include=FALSE}
lifespan_variation <- readRDS("../saves/lifespan_variation_all.rds")
lifespan_mcomparison <- readRDS("../saves/lifespan_mcomparison.rds") %>% mutate_if(is.numeric, round, 2)
lifespan_mgaussianprocess <- readRDS("../saves/lifespan_mgaussianprocess.rds")
lifespan_beast <- readRDS("../saves/lifespan_ibeast.rds")
lifespan_cps <- readRDS("../saves/lifespan_ibeast_cps.rds") %>% round(2) %>% arrange(cp)
lifespan_cps_descr <- readRDS("../saves/lifespan_cps_descr.rds")

@@

High levels of variation pointed to substantial individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]). We found substantial evidence for a non-linear development in gaze following across the lifespan. The Gaussian process model was clearly preferred over the polynomial models due to the highest predictive accuracy according to the LOO ELPD estimates: elpd_diff between Gaussian process and cubic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mcubic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mcubic"]`); elpd_diff between Gaussian process and quadratic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mquadratic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mquadratic"]`); elpd_diff between Gaussian process and linear model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mlinear"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mlinear"]`); all in favor of the Gaussian process model. Moreover, the Gaussian Process model showed the greatest model weight (approximating `r lifespan_mcomparison$modelweight[lifespan_mcomparison$model == "mgaussianprocess"]`). For the imprecision in gaze following, the standard deviation of the Gaussian process (SD = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) indicated nonlinearity.

The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp_mode` major shifts in gaze following during the lifespan (MAP estimate with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp_mode + 1] * 100 %>% round(2)`% probability). The change points occurred at `r lifespan_cps$cp[1]` years (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`], mean imprecision until change point = `r lifespan_cps_descr[1,]$mean`, SD = `r lifespan_cps_descr[1,]$sd`); `r lifespan_cps$cp[2]` years (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`], mean = `r lifespan_cps_descr[2,]$mean`, SD = `r lifespan_cps_descr[2,]$sd`); `r lifespan_cps$cp[3]` years (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`], mean = `r lifespan_cps_descr[3,]$mean`, SD = `r lifespan_cps_descr[3,]$sd`); `r lifespan_cps$cp[4]` years (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`], mean = `r lifespan_cps_descr[4,]$mean`, SD = `r lifespan_cps_descr[4,]$sd`); `r lifespan_cps$cp[5]` years (95% CrI [`r lifespan_cps$lowerCrI[5]`; `r lifespan_cps$upperCrI[5]`], mean = `r lifespan_cps_descr[5,]$mean`, SD = `r lifespan_cps_descr[5,]$sd`); and `r lifespan_cps$cp[6]` years (95% CrI [`r lifespan_cps$lowerCrI[6]`; `r lifespan_cps$upperCrI[6]`], mean = `r lifespan_cps_descr[6,]$mean`, SD = `r lifespan_cps_descr[6,]$sd`). In short: we found a rapid initial improvement in gaze following in early childhood, followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward old age.

@@

Previous studies found that already four-month-olds demonstrate basic gaze following abilities [@delbianco2019developmental]. Since we measured an active location choice on a touchscreen, we could not collect data from infants. In our sample, three-year-olds were still rather imprecise in their gaze following ability (average imprecision was approx. two target widths). How can we explain this divergence? First, we used subtle eye movements as cues. Many existing studies let the agents move eye and head in parallel [@behne2005oneyearolds; @povinelli1997exploitation], establishing a confound with the more salient head movement. Relying exclusively on eye movements might be more difficult for children than presenting them with a combined eye and head orientation [@carpenter1998social]. @silverstein2021infants used a similar manipulation of gaze cues without head rotation and found that 6- to 18-month-olds were around or just above chance for gaze following. The authors argue that infants might fixate on another's face most of the time, while eye movement alone might not be strong enough to guide their attention. Furthermore, our study required participants to (1) precisely follow an agent's gaze, (2) interpret this as a cue, and (3) use the cue to guide their own behavior. It is conceivable that three-year-olds followed the agent's gaze but did not translate this into precise, active behavior. @moll2013primacy argue that social forms of perspective-taking evolve prior to visual perspective-taking, which only emerges within the third year of life. Young children might simply not be interested in a differential, spatial representation of the surrounding objects. Taken together, this might explain why our younger participants located the agent's gaze rather imprecisely.

@@

Our lifespan study showed that gaze following develops throughout childhood, and variation between individuals appears in all age groups. The TANGO has previously been shown to reliably capture inter-individual differences in gaze following [@prein2023tango]. The variation between participants was thus likely genuine and not due to random noise. In Study 2, we aimed to understand the developmental change and individual differences on a process level. We present a theory of gaze following that explains how participants process the available gaze information and trace a line-of-sight to identify the agent's focus. We formalized this inference process in a computational cognitive model that replicates a schematic representation of how participants make inferences in the task’s context (i.e., a model of the task and not the data).

@@

[^6]: In an exploratory analysis, we simulated data for two more alternatives: a line-of-sight tracing model that assumed no inferential noise in the participants' gaze following ability, and a model building up on the line-of-sight tracing with added motor noise. Please note that these did not predict a U-shape pattern, since all target locations would be influenced by the motor noise equally. For further details, see Supplements.

@@

First, we experimentally isolated the vector following component of the TANGO. We designed a new non-social vector following task that shared all crucial design features of the TANGO. Second, we assessed children's social-cognitive abilities by administering a ToM task battery, comprising four tasks from the ToM scale by Wellman and Liu [@wellman2004scaling] and two additional perspective-taking tasks [@flavell1981development; @flavell1981younga]. We reasoned that the TANGO shares task demands with the non-social vector following task while it shares its social context with the ToM tasks. We could, therefore, imagine both an absence or presence of relationship between gaze following and ToM. As stated in our pre-registration, we further assessed whether the two perspective-taking tasks related to gaze following. Our reasoning was that similar underlying mechanisms might be needed to solve these tasks since they require participants to take into account another person's point of view.

@@

We administered four tasks from the @wellman2004scaling ToM scale (see Supplements for further detail). We excluded three tasks: the Diverse Desires task to avoid ceiling effects; and both tasks involving emotions (Belief Emotion and Real-Apparent Emotion), as we aimed at assessing the "cold, cognitive" (vs. "emotional") aspects of social cognition. We added two perspective-taking level-2 tasks ([@flavell1981development; @flavell1981younga]; where children were asked whether a turtle appeared to be on its back or feet / a worm lay on a red or blue blanket from the experimenter's point of view) with the aim of increasing the variability we can capture between individuals, and since we hypothesized that perspective-taking would rely on similar mechanisms than gaze following, both relying on another's person egocentric frame of reference.

@@

(ref:figlab4) **Components of gaze following.** (A) Study procedures. Top: TANGO (i.e., gaze following; social vector following). Bottom: Magnet (i.e., non-social vector following). The left-hand side shows screenshots of the familiarization phase; right-hand side shows screenshots of the test phase. For illustrative purposes, translucent targets are shown in the test phase. Participants cannot see them. (B) Correlations between gaze following, non-social vector following, ToM, and perspective-taking. Dots show the correlation coefficients, error bars represent 95% CIs. Please note that the ToM score acts as an umbrella term and also includes two perspective-taking tasks. (C) Influence of age, perspective-taking and non-social vector following on gaze following. The graph shows the posterior distributions for the respective predictor. Black dots represent means, thicker black lines 80% CrI, and thinner black lines 95% CrI.

@@

In this paper, we have focused on studying variation across ages and individuals. However, our findings rely on participants from Western, Educated, Industrialized, Rich, and Democratic (WEIRD) backgrounds [@henrich2010weirdesta]. Cultural variation has been found in many foundational aspects of cognition and socialization: for example, in parent-child interaction and communication [@nielsen2017persistent]. First evidence suggests that cultural variation in face-to-face interactions does not influence infants' gaze shifts [@hernik2019infant]. While we cannot generalize our here-reported developmental trajectory to different socio-environmental settings, we predict that our presented process-model of gaze following holds true across communities. Analyzing cross-cultural variation and checking for predicted "signature patterns" in the data will inform our modeling work and theory building in further detail [@amir2020crosscultural].

In Study 1, we recruited older participants online which might have selected a particular subgroup of this age range. Seventy-year-olds who have working Wi-Fi connections, know how to use a computer, and are registered on *Prolific* might not be representative of their age group. We can imagine that results from a more diverse, in-person data collection might show different developmental trajectories toward old age.

@@

In three studies, we have illuminated the lifelong development of precisely estimating another's gaze direction, and the psycho-physical process behind this. We have shown that gaze following continues to develop beyond infancy, and that individuals differ in their precision levels to localize the gaze direction of an agent. Our proposed process-level theory of gaze following modeled individual differences in precision as varying levels of uncertainty in the estimated gaze vectors. Consequently, we found that imprecision in gaze following relates to non-social vector following, as proposed by the model. Additionally, gaze following was linked to visual perspective-taking but no other aspects of ToM. The present research shows how precise and reliable measures and process models jointly inform each other and lead to a more comprehensive understanding of the psychological phenomenon in question.

### 2024-04-17 stats/lifespan.Rmd
ready for revision

>>>>

```{r changepoints_descr}
cps <- readRDS("../saves/lifespan_ibeast_cps.rds") %>% 
  arrange(cp)

cps_descr <- mdata %>% 
  mutate(
   age_cp = case_when(
     age < cps$cp[1] ~ "1",
     age < cps$cp[2] ~ "2",
     age < cps$cp[3] ~ "3",
     age < cps$cp[4] ~ "4",
     age < cps$cp[5] ~ "5",
     age < cps$cp[6] ~ "6",
     age > cps$cp[6] ~ "7",
   ) 
  ) %>% 
  group_by(age_cp) %>%
  summarize(
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    n = n(),
    mean = mean(mean_imprecision, na.rm = T) %>% round(2),
    sd = sd(mean_imprecision, na.rm = T) %>% round(2),
  )

saveRDS(cps_descr, "../saves/lifespan_cps_descr.rds")
```
